---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dqa4mortality: Mortality Data Quality Assessment Framework

<!-- badges: start -->
<!-- badges: end -->

The National Vital Statistics System (NVSS) is managed by the U.S. Centers for Disease Control and
Prevention and is the system of record for all vital events (e.g., births, deaths) in the United
States. This system, for example, is used to manage approximately 2.8 million death records annually
and was constructed to ensure accurate and timely data availability (NCHS 2024). Death certificate
data, however, have at times been found to incomplete, inaccurate, and untimely, affecting the
utility of the death certificate data
([Flagg 2021](https://stacks.cdc.gov/view/cdc/100414)).
To address some of these challenges, over the
last decade, 54 of 57 jurisdictions implemented an electronic death reporting system (EDRS). These
EDRSs allow for electronic capture of data, thus gaining efficiency in data sharing and reporting.

Although the use of EDRSs, especially with embedded edit check tools (e.g.,
[Validations and Interactive Edits Web Service](https://www.cdc.gov/nchs/data/nvss/modernization/VIEWS-Technical-User-Info-508.pdf)),
has improved data quality,
[challenges persist (Flagg 2021)](https://stacks.cdc.gov/view/cdc/100414).
Comprehensive, monitoring of data quality over time is essential to identify gaps and identify
opportunities to improve data quality.

dqa4mortality implements functions to execute Data Quality Assessment
Framework fort mortality data.

All code in this section of the repository is intended for R users. If you prefer to use Python, please see [../../python/dqa4mortality](python/dqa4mortality).

## Installation

You can install dqa4mortality from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nightingaleproject/Mortality-Data-Quality-Assessment-Examples/R/dqa4mortality")
```

## Data Requirements

Data used for metric calculation should have a format where each row corresponds to one death record and each column corresponds to an attribute for the death record, such as "Date Certified" or "Underlying Cause of Death". ICD codes included in the data should not include periods (".").

You can load your CSV data by using the following function, replacing "file_path" with the path to your mortality CSV:
```{r, eval=FALSE}
library(dqa4mortality)
death_records <- load_death_records(file_path)
```


Example data is included in this package, called *synthetic_death_records*. To look at the first few rows and columns of this example data:

```{r}
library(dqa4mortality)
head(synthetic_death_records[,1:33])
```


The raw CSV is also included in this repository; please see [SyntheticDeathRecordData.csv](data/SyntheticDeathRecordData.csv).

## Functions

dqa4mortality includes several functions to assess quality of your given mortality data. Functions beginning with "proportion\*" calculate metrics across the whole dataset, whereas those beginning with "certifier_proportion\*" calculate metrics by certifier within the dataset.

Functions to assess data quality are:

  * (proportion/certifier_proportion)_not_certified_within_required_period() – Proportion of records that were not certified within the expected time – jurisdictions typically specify how quickly a death should be certified after the actual date of death.

  * (proportion/certifier_proportion)_with_incomplete_funeral_director_fields() – Proportion of records with at least one “funeral director” field incomplete – funeral directors are typically responsible for providing identifying information and some demographic information.

  * (proportion/certifier_proportion)_with_incomplete_medical_certifier_fields() – Proportion of records with at least one “medical certifier” field incomplete – the medical certifier is the medical professional or authorized person who determines the cause of death and manner of death.

  * (proportion/certifier_proportion)_with_incomplete_demographic()* – Proportion of records with incomplete information for any of several demographic fields – demographic fields include variables such as age, race/ethnicity, and occupation.

  * (proportion/certifier_proportion)_with_one_cause() – Proportion of records with only one cause of death condition – multiple clinical conditions are typically reported by the medical certifier within Part I and Part II of the death certificate.

  * (proportion/certifier_proportion)_with_unsuitable_underlying() – Proportion of records with unsuitable underlying cause of death (UCOD) – unsuitable UCODs are those which are unknown and ill-defined; immediate and intermediate; and nonspecific.
  
For more information on inputs, outputs, and return values for each function, see the help text for each function by prefacing it with a question mark in your console. For example:

```r
?proportion_not_certified_within_required_period
```

## Usage

Let's start with assessing a simple metric on our synthetic data, proportion with unsuitable underlying causes of death: 

```{r example}
library(dqa4mortality)

prop <- proportion_with_unsuitable_underlying(
  synthetic_death_records,
  underlying_cause_of_death_column = "Underlying COD"
)
```
We can see that it printed out the rounded proportion of records with an unsuitable underlying cause of death. Additionally, if we look at the value that it stored, `prop`:

```{r}
print(prop)
```

We can see that it stored the exact proportion in our variable. This can be useful for noting to examine small changes in proportions over time.

Additionally, we can examine that same metric by certifier:

```{r}
certifier_prop <- certifier_proportion_with_unsuitable_underlying(
  synthetic_death_records,
  underlying_cause_of_death_column = "Underlying COD",
  certifier_name_column = "Certifier Name"
)
```


```{r}
head(certifier_prop)
```

We can see that our proportion is printed by Certifier Name in our dataset, and is sorted by the certifier with the highest proportion of unsuitable underlying causes of death.

We could then use the data frame that's returned to visualize our certifiers with the highest proportions:

```{r, fig.height=3, fig.width=2}
library(ggplot2)

ggplot(
  certifier_prop[1:5,], 
  aes(factor(`Certifier Name`, levels = `Certifier Name`), Proportion)
)+
  geom_col(color = "black", fill = "lightblue")+
  theme_bw()+
  scale_x_discrete(guide = guide_axis(angle = 45))+
  xlab("Certifier Name")
```

## Disclosure

Artificial Intelligence (AI) software was consulted in the development of this code and to ensure that R and Python versions of the code run identically.

## License

Copyright 2024 The MITRE Corporation

Licensed under the Apache License, Version 2.0 (the "License"); you may not use this file except in compliance with the License. You may obtain a copy of the License at

```
http://www.apache.org/licenses/LICENSE-2.0
```

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS, WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied. See the License for the specific language governing permissions and limitations under the License.

### Contact Information

For questions or comments about this code repository please send email to

    nvssmodernization@cdc.gov
