---
title: "Calculating Metrics for Assessing Mortality Data Quality"
author: "MITRE"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following code demonstrates some simple examples of calculating metrics that are part of the Data Quality Assessment Framework and  Jurisdictional Playbook.

## Calculating Metrics Related to Unsuitable Underlying Cause of Death

The following code demonstrates how mortality records can be evaluated against a list of unsuitable underlying cause of death codes, derived from **R/proportion_with_unsuitable_underlying.R**.

We start by loading necessary libraries for processing, as well as our data:

```{r}
# Load necessary libraries
library(here)

# Load supporting functions
source(here::here("code", "R", "dqaf_metrics.R"))

# Load the death records data
file_path <- here::here("data", "SyntheticDeathRecordData.csv")
death_records <- 
  load_death_records(file_path)
```

Then we calculate which causes of death in our data are unsuitable:

```{r}
# Load the unsuitable causes of death data
unsuitable_causes <- read.csv(here::here("data", "unsuitable_COD_codes.csv"))

# Extract the unsuitable codes
unsuitable_codes <- unsuitable_causes$code

# Create a new column that is TRUE when the underlying COD is unsuitable
death_records[, "Unsuitable Underlying"] <- sapply(
  death_records$`Underlying COD`, 
  function(code){
    return(any(startsWith(code, unsuitable_codes)))
  }
)
```

How many records have an unsuitable cause of death?

```{r}
# Calculate the proportion of records with an unsuitable underlying cause of death
proportion <- calculate_proportion(
  death_records, 
  metric = "Unsuitable Underlying",
  metric_description = "unsuitable underlying cause of death"
)
```

Let's examine the records with unsuitable cause of death by certifier:

```{r}
# Group the records by certifier and calculate the proportion of unsuitable records for each certifier
certifier_proportions <- calculate_proportion_by_column(
  death_records, 
  metric = "Unsuitable Underlying",
  metric_description = "unsuitable underlying cause of death",
  column = "Certifier Name"
)
```

Let's visualize that using ggplot to see the certifiers with the highest proportion of unsuitable records:

```{r, out.height="120%"}
library(ggplot2)

ggplot(
  certifier_proportions[order(certifier_proportions$Proportion, 
                              decreasing = FALSE),], 
       aes(
         factor(.data[["Certifier Name"]], levels = .data[["Certifier Name"]]),
         .data[["Proportion"]]))+
  geom_col(fill = "lightblue", color = "black")+
  xlab("Certifier Name")+
  coord_flip()+
  theme_bw()
```

