---
title: "Calculating Metrics for Assessing Mortality Data Quality"
author: "MITRE"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following code demonstrates some simple examples of calculating metrics that are part of the Data Quality Assessment Framework and  Jurisdictional Playbook.

## Calculating Metrics Related to Unsuitable Underlying Cause of Death

The following code demonstrates how mortality records can be evaluated against a list of unsuitable underlying cause of death codes, derived from **R/proportion_with_unsuitable_underlying.R**.

We start by loading necessary libraries for processing, as well as our data:

```{r}
library(here)

# Specify the relative path to the data location
data_path <- here::here("data")

# Load the death records data
death_records <- read.csv(
  file.path(data_path, "SyntheticDeathRecordData.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE,
  na.strings = ""
)
```

Then we calculate which causes of death in our data are unsuitable:

```{r}
# Load the unsuitable causes of death data
unsuitable_causes <- read.csv(file.path(data_path, "unsuitable_COD_codes.csv"))

# Extract the unsuitable codes
unsuitable_codes <- unsuitable_causes$code

# Create a new column that is TRUE when the underlying COD is unsuitable
death_records[, "Unsuitable Underlying"] <- sapply(
  death_records$`Underlying COD`, 
  function(code){
    return(any(startsWith(code, unsuitable_codes)))
  }
)
```

How many records have an unsuitable cause of death?

```{r}
# Calculate the proportion of records with an unsuitable underlying cause of death
proportion <- mean(death_records$`Unsuitable Underlying`, na.rm = TRUE)

cat(paste("The proportion of records with an unsuitable underlying cause of death is",
          round(proportion, 2),
          "\n"))
```

Let's examine the records with unsuitable cause of death by certifier:

```{r}
# Group the records by certifier and calculate the proportion of unsuitable records for each certifier
certifier_proportions <- 
  aggregate(
    list("Proportion" = death_records$`Unsuitable Underlying`),
    list("Certifier Name" = death_records$`Certifier Name`),
    FUN = mean,
    na.rm = TRUE)

# Print the proportions for each certifier
for(i in 1:nrow(certifier_proportions)) {
  cat(paste("The proportion of records with an unsuitable underlying cause of death provided by",
            certifier_proportions[i, "Certifier Name"],
            "is",
            round(certifier_proportions[i, "Proportion"], 2), "\n"))
}
```

Let's visualize that using ggplot to see the certifiers with the highest proportion of unsuitable records:

```{r, out.height="120%"}
library(ggplot2)

ggplot(
  certifier_proportions[order(certifier_proportions$Proportion, 
                              decreasing = FALSE),], 
       aes(
         factor(.data[["Certifier Name"]], levels = .data[["Certifier Name"]]),
         .data[["Proportion"]]))+
  geom_col(fill = "lightblue", color = "black")+
  xlab("Certifier Name")+
  coord_flip()+
  theme_bw()
```

